Testing Statistical Graphics

### 2022-06-27 index.qmd
initial read through at face value with minor edits.

---
title: Testing Statistical Graphics
author-list:
  - name: Emily Robinson
    dept: Statistics Department
    affiliation: CalPoly
    email: earobinson@huskers.unl.edu
  - name: Heike Hofmann
    dept: Statistics Department
    affiliation: Iowa State University
    email: hofmann@iastate.edu
  - name: Susan Vanderplas
    dept: Statistics Department
    affiliation: University of Nebraska, Lincoln
    email: svanderplas2@unl.edu

@@

# Testing Methods and Engagement
<!-- Lineups, You Draw It, Direct Assessment -->

@@

Statistical lineups\cite{bujaStatisticalInferenceExploratory2009,wickhamGraphicalInferenceInfovis2010} are a more generally applicable testing tool for perceptual questions such as "which chart displays this data more clearly"\cite{power}. 
As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise; if viewers consistently pick the target plot instead of any of the 19 null plots, then the target plot is said to be visually significant\cite{majumderValidationVisualStatistical2013,vanderplasStatisticalSignificanceCalculations2021} and a "see" value (the visual analogue of a p-value for a statistical test) can be calculated\cite{niladriroychowdhurySeeValueApp2020}.
In another variation of statistical lineup procedure, two models are compared, with target plots from each model embedded in the array of 20 total plots; the remaining null plots are constructed from a mixture model that blends the two competing models. This variation allows the experimenter to assess graphical design choices to determine whether they effectively emphasize structural differences in the data\cite{clusters}.
Statistical lineups do not require the experimenter to provide any contextual information for the data: all of the necessary information is embedded in the choice of a null statistical model, which is convenient for testing purposes, but does not require engagement at a level that would allow for testing the viewer's understanding of the information shown in the chart. 
An example of a statistical lineup for testing the perception of exponential growth is shown in @fig-lineup-log.

@@

![This question asks users to write out a description of how the population of Ewoks changes over time, without any further cues, to determine whether participants default to multiplicative or additive descriptions.](Estimation_describe_plot_crop.png){#fig-estimation-describe}

@@

![This question asks participants for a numerical estimate, but provides a basic calculator and scratchpad. All user interactions with the calculator and scratchpad are logged, providing insight into the user's thought process.](Estimation_numerical_screenshot_crop.png){#fig-estimation-calc}

In our experience, one of the most difficult components of designing an experiment which asks users to directly estimate information from a chart using a full scenario (background information, etc. as well as contextual details from the chart) is that the wording of such questions must be extremely carefully determined, and even when great care is taken with the construction of the question, participant answer accuracy is fundamentally limited by the fact that many participants do not read and interpret the question with the care and precision that it was written.
Questions that ask participants to e.g. estimate the multiplicative difference in a quantity at two timepoints may be misunderstood as asking for an estimate of the additive difference, and the resulting estimates are then one or more orders of magnitude off of the correct answer.
This is one area where lineup tests are convenient - they do not depend on participants to understand the nuances of language or scenarios built around the chart under investigation.
However, in some situations it may be sufficient to ask participants to estimate direct numerical quantities that have little contextual information, as we did in \cite{vanderplasFramedReproducingRevisiting2019} when assessing the accuracy of framed plots re-created from the Statistical Atlas.

@@

We cannot overstate the importance of visual assessment of your model stimuli, preferrably with fresh eyes. We highly recommend performing a series of think-aloud pilot testing before deploying an experiment. In support of this assessment, we offer up a cautionary tale of our own experience: that of \cite{clusters}, where we designed an experiment to test which plot aesthetics promoted discovery of linear trends and/or clusters.

@@

The primary human design factor to be aware of is that visual tasks and assessing statistical graphics can be extremely cognitively taxing. In our experience, it is difficult to expect participants to evaluate more than about 15 statistical lineups in one sitting. It is easy to understand how this arises when we consider the concept of decision fatigue: each standard lineup has 20 panels, so there are 190 pairwise comparisons to evaluate between the plots of just a single lineup (though it is unlikely that participants actually sequentially evaluate each comparison in a combinatorial fashion). Tasks which are more interactive, such as you-draw-it, may be somewhat easier for participants, but it is unlikely that participants would be willing to complete more than about 20 tasks in one sitting even with tasks that require fewer decisions and more engagement. At some point, participants' interest in accuracy will decline, and the researcher is better off fitting the task within the window where participants are attentive and engaged.

It can be extremely helpful to include "practice" demonstrations of the task to show the basic process, logic, and reasoning. While it is tempting to make these tasks fully representative of the type of judgement which will be required of participants, practice tasks which are too close to the experimental task may bias participants; we have found that it works well to have a relatively easy practice task which utilizes a slightly different type of plot and/or type of data than what will be tested in the experiment. In cases which require interactivity, gif animations of the task being carried out are useful, as are additional visual cues, such as the yellow box used in the You-Draw-It task[^1] to indicate that there were points which were not completed.

@@

We have conducted visualization experiments using a wide variety of tools: custom web servers running experiments using PHP black magic, Qualtrics surveys for static graphics, and writing full Shiny applications that control every part of the experimental process from instructions to providing completion codes for payment. 
In our research, Shiny has provided the right balance between control over the experimental setting, procedure, etc. and the intricate details of web server management, however, this balance is likely different for every group and potentially for every experiment. 
Likewise, we have recruited participants using Amazon Mechanical Turk, Prolific, Reddit and other social media sites, and email; each recruitment method has trade-offs between cost, convenience, and control over demographic variables.

[^1]: See a gif of testing with You Draw It [here](https://i.imgur.com/GM5YSen.gif)

@@

**Direct Interactions** If participants are making predictions and/or fitting visual statistics, we have had success analyzing these responses by comparing the responses to results from a statistical model to determine how visual statistics differ from the numerical quantities derived mathematically. For instance, we calculate the deviation between participant responses and the linear regression in You-Draw-It experiments, then fitted generalized additive mixed models to summarize the results across different experimental conditions to assess how user-drawn predictions deviated from the statistical estimates. 
In other direct interactions, it may be useful to compare participant selections or annotations to closest points on the chart to assess anchoring behavior; for discrete selections, methods discussed in numerical estimation may also be useful.

>>>>

---
title: Testing Statistical Graphics
author-list:
  - name: Emily Robinson
    dept: Statistics Department
    affiliation: California Polytechnic State University
    email: erobin17@calpoly.edu
  - name: Heike Hofmann
    dept: Statistics Department
    affiliation: Iowa State University
    email: hofmann@iastate.edu
  - name: Susan Vanderplas
    dept: Statistics Department
    affiliation: University of Nebraska, Lincoln
    email: svanderplas2@unl.edu

@@

# Testing Methods and \ear{User} Engagement
<!-- Lineups, You Draw It, Direct Assessment -->

@@

Statistical lineups\cite{bujaStatisticalInferenceExploratory2009,wickhamGraphicalInferenceInfovis2010} are a more generally applicable testing tool for perceptual questions such as "which chart displays this data more clearly"\cite{power}. 
As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise; if viewers consistently pick the target plot instead of any of the 19 null plots, then the target plot is said to be visually significant\cite{majumderValidationVisualStatistical2013,vanderplasStatisticalSignificanceCalculations2021} and a "see" value (the visual analogue of a p-value for a statistical test) can be calculated\cite{niladriroychowdhurySeeValueApp2020}.
In another variation of the statistical lineup procedure, two models are compared, with target plots from each model embedded in the array of 20 total plots; the remaining null plots are constructed from a mixture model that blends the two competing models. This variation allows the experimenter to assess graphical design choices to determine whether they effectively emphasize structural differences in the data\cite{clusters}.
Statistical lineups do not require the experimenter to provide any contextual information for the data: all of the necessary information is embedded in the choice of a null statistical model, which is convenient for testing purposes, but does not require engagement at a level that would allow for testing the viewer's understanding of the information shown in the chart. 
An example of a statistical lineup for testing the perception of exponential growth is shown in @fig-lineup-log.

@@

![This question asks users to write out a description of how the population of Ewoks changes over time, without any further cues, to determine whether participants default to multiplicative or additive \ear{language} descriptions.](Estimation_describe_plot_crop.png){#fig-estimation-describe}

@@

![This question asks participants for a numerical estimate, but provides a basic calculator and scratchpad. All user interactions with the calculator and scratchpad are logged, providing insight into the user's thought process \ear{and estimation strategy}.](Estimation_numerical_screenshot_crop.png){#fig-estimation-calc}

In our experience, one of the most difficult components of designing an experiment which asks users to directly estimate information from a chart using a full scenario (background information, etc. as well as contextual details from the chart) is that the wording of such questions must be extremely carefully determined, and even when great care is taken with the construction of the question, participant answer accuracy is fundamentally limited by the fact that many participants do not read and interpret the question with the care and precision that it was written.
Questions that ask participants to e.g. estimate the multiplicative difference\ear{change} in a quantity at two time points may be misunderstood as asking for an estimate of the additive difference, and the resulting estimates are then one or more orders of magnitude off of the correct answer.
This is one area where lineup tests are convenient - they do not depend on participants to understand the nuances of language or scenarios built around the chart under investigation.
However, in some situations it may be sufficient to ask participants to estimate direct numerical quantities that have little contextual information, as we did in \cite{vanderplasFramedReproducingRevisiting2019} when assessing the accuracy of framed plots re-created from the Statistical Atlas.

@@

We cannot overstate the importance of visual assessment of your model stimuli, preferably with fresh eyes. We highly recommend performing a series of think-aloud pilot testing before deploying an experiment. In support of this assessment, we offer up a cautionary tale of our own experience: that of \cite{clusters}, where we designed an experiment to test which plot aesthetics promoted discovery of linear trends and/or clusters.

@@

The primary human design factor to be aware of is that visual tasks and assessing statistical graphics can be extremely cognitively taxing. In our experience, it is difficult to expect participants to evaluate more than about 15 statistical lineups in one sitting. It is easy to understand how this arises when we consider the concept of decision fatigue: each standard lineup has 20 panels, so there are 190 pairwise comparisons to evaluate between the plots of just a single lineup (though it is unlikely that participants actually sequentially evaluate each comparison in a combinatorial fashion). Tasks which are more interactive, such as \'You Draw It'}you-draw-it, may be somewhat easier for participants, but it is unlikely that participants would be willing to complete more than about 20 tasks in one sitting even with tasks that require fewer decisions and more engagement. At some point, participants' interest in accuracy will decline, and the researcher is better off fitting the task within the window where participants are attentive and engaged.

It can be extremely helpful to include "practice" demonstrations of the task to show the basic process, logic, and reasoning. While it is tempting to make these tasks fully representative of the type of judgement which will be required of participants, practice tasks which are too close to the experimental task may bias participants; we have found that it works well to have a relatively easy practice task which utilizes a slightly different type of plot and/or type of data than what will be tested in the experiment. In cases which require interactivity, gif animations of the task being carried out are useful, as are additional visual cues, such as the yellow box used in the \ear{'You Draw It'} task[^1] to indicate that there were points which were not completed.

@@

We have conducted visualization experiments using a wide variety of tools: custom web servers running experiments using PHP \ear{explain acronym} black magic, Qualtrics surveys for static graphics, and writing full Shiny applications that control every part of the experimental process from instructions to providing completion codes for payment. 
In our research, Shiny has provided the right balance between control over the experimental setting, procedure, etc. and the intricate details of web server management, however, this balance is likely different for every group and potentially for every experiment. 
Likewise, we have recruited participants using Amazon Mechanical Turk, Prolific, Reddit and other social media sites, and email; each recruitment method has trade-offs between cost, convenience, and control over demographic variables.

[^1]: See a gif of testing with 'You Draw It' [here](https://i.imgur.com/GM5YSen.gif)

@@

**Direct Interactions** If participants are making predictions and/or fitting visual statistics, we have had success analyzing these responses by comparing the responses to results from a statistical model to determine how visual statistics differ from the numerical quantities derived mathematically. For instance, we calculate the deviation between participant responses and the linear regression in 'You Draw It' experiments, then fitted generalized additive mixed models to summarize the results across different experimental conditions to assess how user-drawn predictions deviated from the statistical estimates. 
In other direct interactions, it may be useful to compare participant selections or annotations to closest points on the chart to assess anchoring behavior; for discrete selections, methods discussed in numerical estimation may also be useful.
